---
title: "Nonlinear Curve Fitting"
author: "Krzysztof Trajkowski"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
               toc: yes
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(fig.align="center",
               fig.width=6, fig.height=4,
               dev.args=list(pointsize=8),
               par=TRUE)

knit_hooks$set(par = function(before, options, envir)
  { if(before && options$fig.show != "none") 
       par(mar=c(4,4,1,1)+0.1, mgp=c(3,0.6,0),las=1)
})
```

# Library

```{r}
library("FitCurve")
```

# Nonlinear Least Squares

## Exponential curve

Exponential function - **SSexpo**:
$$f(x)=ab^x$$

### Levenberg-Marquardt algorithm

```{r}
expo <- nlsLM(USA~SSexpo(year,a,b),
              data=subset(smoking,year<="1949"))
summary(expo)
```

### Plot

```{r}
with(plot(y=USA,x=year,type="l",lwd=2,
          xlim=c(1900,2030),ylim=c(0,12)),
     data=subset(smoking,year<="1949"))
title("Sales of cigarettes per adult per day in USA 1927-1949")
p <- coef(expo)
curve(SSexpo(x,p[1],p[2]),add=TRUE,col="orange3",lwd=2)
```

## Logistic curve

Logistic function - [SSlogis](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/SSlogis.html):
$$f(x)=\frac{Asym}{1+\exp{\left(\frac{xmid-x}{scal}\right)}}$$

* Inflection point:

$$xmid$$

Logistic function - **SSlogis1**:
$$a=Asym,\quad b=\frac{xmid}{scal},\quad c=-\frac{1}{scal}$$
$$f(x)=\frac{a}{1+\exp{(b+cx)}}$$

* Inflection point:

$$-b/c$$

Logistic function - **SSlogis2**:
$$\alpha=Asym,\quad \beta=\exp{\left(\frac{xmid}{scal}\right)},\quad \gamma=\frac{1}{scal}$$
$$f(x)=\frac{\alpha}{1+\beta\exp{(-\gamma x)}}$$

* Inflection point:

$$\ln(\beta)/\gamma$$

and other models of the [drc](https://cran.r-project.org/web/packages/drc/) package.

### nl2sol algorithm

```{r}
logi <- nls(USA~SSlogis(year,Asym,xmid,scal),
            algorithm = "port",
            data=subset(smoking,year<="1966"))
summary(logi)
```

### Plot

```{r}
with(plot(y=USA,x=year,type="l",lwd=2,
          xlim=c(1900,2030),ylim=c(0,12)),
     data=subset(smoking,year<="1966"))
title("Sales of cigarettes per adult per day in USA 1927-1966")
p <- coef(logi)
curve(SSlogis(x,p[1],p[2],p[3]),add=TRUE,col="orange3",lwd=2)
```

## Bell curve

Amplitude version of Gaussian peak function - **SSgaussAmp**:
$$f(x)=y_{0}+A\exp{\left(\frac{-(x-x_c)^2}{2w^2}\right)}$$

* Inflection point:

$$x_L=x_c-\frac{1}{w},\quad x_U=x_c+\frac{1}{w}$$

Area version of Gaussian Function - **SSgaussAre**:
$$f(x)=y_0+\frac{A}{w\sqrt{\pi/2}}\exp{\left(-2\frac{(x-x_c)^2}{w^2}\right)}$$

* Inflection point:

$$x_L=x_c-\frac{w}{2},\quad x_U=x_c+\frac{w}{2}$$

### Gauss-Newton algorithm

```{r}
amp <- nls(USA ~ SSgaussAmp(year,y0,A,xc,w),
           data=smoking)
summary(amp)
```

### Plot

```{r,fig.height=4,fig.width=6}
with(plot(year,USA,type="l",lwd=2,
          xlim=c(1900,2030),ylim=c(0,12)),
     data=smoking)
title("Sales of cigarettes per adult per day in USA 1927-2010")
p <- coef(amp)
curve(SSgaussAmp(x,p[1],p[2],p[3],p[4]),add=TRUE,col="orange3",lwd=2)
```

# Maximum Likelihood Estimation

## Four-Parameter Logistic curve

Four-Parameter Logistic -  [SSfpl](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/SSfpl.html):
$$f(x)=\frac{B-A}{1+\exp{\left(\frac{xmid-x}{scal}\right)}}+A$$

### L-BFGS-B algorithm

```{r}
p <- getInitial(population~SSfpl(year,A,B,xmid,scal),data=population)
```
```{r}
fpl_nor <- mle2(population~dnorm(mean= SSfpl(year,A,B,xmid,scal),sd= sd),
                start=list(A= p[1], B= p[2], xmid= p[3], scal= p[4], sd= 1),
                method= "L-BFGS-B", data= population,
                lower=c(A=-Inf,B=-Inf,xmid=-Inf,scal=-Inf,sd=0))
summary(fpl_nor)
```

### Plot

```{r,fig.height=4,fig.width=6}
with(plot(year,population,type="l",lwd=3,
          xlim=c(1800,2100),ylim=c(0,45e+07)),
     data=population)
title("Population in USA 1790-2010")
p <- coef(fpl_nor)
curve(SSfpl(x,p[1],p[2],p[3],p[4]),add=TRUE,col="orange3",lwd=1.5)
```

## Logistic curve

Logistic function - [SSlogis](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/SSlogis.html):
$$f(x)=\frac{Asym}{1+\exp{\left(\frac{xmid-x}{scal}\right)}}$$

### L-BFGS-B algorithm

```{r}
p <- getInitial(population~SSlogis(year,Asym,xmid,scal),data=population)
```
```{r}
logi_nor <- mle2(population~dnorm(mean= SSlogis(year,Asym,xmid,scal),sd= sd),
                 start=list(Asym= p[1], xmid= p[2], scal= p[3], sd= 1),
                 method= "L-BFGS-B", data= population,
                 lower=c(Asym=-Inf,xmid=-Inf,scal=-Inf,sd=0))
summary(logi_nor)
```

### Plot

```{r,fig.height=4,fig.width=6}
with(plot(year,population,type="l",lwd=3,
          xlim=c(1800,2100),ylim=c(0,45e+07)),
     data=population)
title("Population in USA 1790-2010")
p <- coef(logi_nor)
curve(SSlogis(x,p[1],p[2],p[3]),add=TRUE,col="orange3",lwd=1.5)
```

# A list of all functions

---

**SSexpo** - exponential model:
$$f(x)=ab^x$$

---

**SSexpoGen**	- general exponential model:
$$f(x)=a\exp{(bx)}$$

---

**SSexpoInv**	- exponential inverse model:
$$f(x)=a\exp{\left(b\frac{1}{x}\right)}$$

---

**SSgaussAmp** - amplitude of gaussian model:
$$f(x)=y_{0}+A\exp{\left(\frac{-(x-x_c)^2}{2w^2}\right)}$$

---

**SSgaussAre** - area of gaussian model:
$$f(x)=y_0+\frac{A}{w\sqrt{\pi/2}}\exp{\left(-2\frac{(x-x_c)^2}{w^2}\right)}$$

---

**SShyper**	- hyperbolic model:
$$f(x)=\frac{ax^2}{x+b}$$

---

**SSlogis1** - logistic model:
$$f(x)=\frac{a}{1+\exp{(b+cx)}}$$

----

**SSlogis2** - logistic model:
$$f(x)=\frac{\alpha}{1+\beta\exp{(-\gamma x)}}$$

---

**SSlorentz** - lorentz model:
$$f(x)=y_0+A\frac{2w}{\pi4(x-x_c)^2+w^2}$$

---

**SSparLog** - parabola logarithmic model:
$$f(x)=ax^{b+c\ln(x)}$$

---

**SSpower** - power model:
$$f(x)=ax^b$$

---

**SSpowExpo** - power-exponential model:
$$f(x)=ax^b\exp{(cx)}$$

---

**SSpowExpoInv** - power-exponential-inverse model:
$$f(x)=ax^b\exp{\left(c\frac{1}{x}\right)}$$

---

**SSpsVoigt1** - Pseudo-Voigt model:
$$f(x)=y_0+A\left[\nu\frac{2w}{\pi4(x-x_c)^2+w^2}+(1-\nu)\frac{\sqrt{4\ln 2}}{\sqrt{\pi}w}\exp\left(-\frac{4\ln 2}{w^2}(x-x_c)^2\right)\right]$$

---

**SSpsVoigt2** - Pseudo-Voigt model with different FWHM ($w_L$ and $w_G$):
$$f(x)=y_0+A\left[\nu\frac{2w_L}{\pi4(x-x_c)^2+w_L^2}+(1-\nu)\frac{\sqrt{4\ln 2}}{\sqrt{\pi}w_G}\exp\left(-\frac{4\ln 2}{w_G^2}(x-x_c)^2\right)\right]$$

---

**SStorn1** - Tornquist I model:
$$f(x)=\frac{ax}{x+b}$$

---

**SStorn2** - Tornquist II model:
$$f(x)=\frac{a(x-c)}{x+b}$$

---

**SStorn3** - Tornquist III model:
$$f(x)=\frac{ax(x-c)}{x+b}$$

---

**SSworking** - Working model:
$$f(x)=\exp{\left[a+b\left(\frac{1}{x}\right)\right]}$$

---

